Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-05-15 talk/dls2014/report/report.tex
move report's content to the paper

% Title.
% ------
\title{C7: Fast software transactional memory for dynamic languages}
%
% Single address.
% ---------------
%\name{Markus P\"uschel\thanks{The author thanks Jelena Kovacevic. This paper
%is a modified version of the template she used in her class.}}
%\address{Department of Computer Science\\ ETH Z\"urich\\Z\"urich, Switzerland}

@@

 cfbolz{citation again needed for the whole subsection}

@@

In this section, we describe the general model of our TM system. This
should clarify the general semantics using commonly used terms from
the literature.

@@

As required for TM systems, we guarantee complete \emph{isolation
}and \emph{atomicity} for transactions at all times. Furthermore,
the isolation provides full \emph{opacity} to always guarantee a consistent
read set.

\cfbolz{this paragraph is hard to understand without giving an example (eg console printing) when it is useful}

We support the notion of \emph{inevitable transactions} that are always
guaranteed to commit. There is always at most one such transaction
running in the system. We use this kind of transaction to provide
\emph{strong isolation} by running non-transactional code in the context
of inevitable transactions and to still provide the \emph{serializability}
of all transaction schedules.

@@

To get references to objects that are valid in all threads, we will
use \cfbolz{use for what?} the object's offset inside the segment. Since all segments are
copies of each other, the \emph{Segment Offset (SO)} will point to the
private version of an object in all threads/segments. To then
translate this SO to a real virtual memory address when used inside a
thread, we need to add the thread's segment start address to the
SO. The result of this operation is called a \emph{Linear Address
  (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.

\cfbolz{here it needs to say that this is x86 specific}

To make this address translation efficient, we use the segment
register $\%gs$. When this register points to a thread's segment start
address, we can instruct the CPU to perform the above translation from
a reference of the form $\%gs{::}SO$ to the right LA on its own.

@@

To provide atomicity for a transaction, we want to make changes globally
visible on commit. We also need to be able to completely abort a
transaction without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronize all threads
  so that all of them are waiting in a safe point. In the committing
  transaction, we go through all objects in the write set and check if
  another transaction in a different segment read the same object.
  Conflicts are resolved again by either the committing or the other
  transaction waiting or aborting.\\
  We then push all changes of modified objects in private pages to all
  the pages in other segments, including the sharing-segment (segment
  0). \cfbolz{can it really happen that you push pages to other segments? I thought it's always just back to the sharing segment}
\item [{Abort:}] On abort the transaction will forget about all the
  changes it has done. All objects in the write set are reset by
  copying their previous version from the sharing-segment into the
  private pages of the aborting transaction.
  \cfbolz{why doing any copying? aren't the pages re-shared instead?}
\end{description}

@@

Garbage collection plays a big role in our TM system. The GC is
generational and has two generations.

\cfbolz{maybe use "young" and "old" generation, if there are only two}

The \textbf{first generation}, where objects are considered to be
\emph{young} and reside in the \emph{Nursery}, is collected by
\emph{minor collections}. These collections move the surviving objects
out of the nursery into the old object space, which can be done
without stopping other threads. This is done either if the nursery has
no space left anymore or if we are committing the current
transaction. Consequently, all objects are old and the nursery empty
after a transaction commits.  Furthermore, all objects in the nursery
were always created in the current transaction. This fact is useful
since we do not need to call any barrier on this kind of objects.

@@

The \textbf{second generation}, where objects are considered to be
\emph{old} and never move again, is collected by \emph{major
  collections}.  These collections are implemented in a stop-the-world
kind of way and first force minor collections in all threads. The
major goal is to free objects in the old objects space. Furthermore,
we optimistically re-share pages that do not need to be private
anymore.

@@

Committing a transaction needs a bit more work. First, we synchronize
all threads so that the committing one is the only one running and all
the others are waiting in a safe point. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads/segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit. \cfbolz{why does waiting help?}

After verifying that there are no conflicts anymore, we copy all our
changes done to the objects in the write set to all other segments,
including the sharing-segment. This is safe since we synchronized all
threads. We also need to push overflow objects generated by minor
collections to other segments, since they may reside partially in
private pages. At that point we also get a new
\lstinline!overflow_number!  by increasing a global one, so that it
stays globally unique for each transaction. Increasing the
\lstinline!read_version!  is then done at the start of a new
transaction.

@@

>>>>

% Title.
% ------
\title{Virtual Memory Assisted Transactional Memory for Dynamic Languages}
%
% Single address.
% ---------------
%\name{Markus P\"uschel\thanks{The author thanks Jelena Kovacevic. This paper
%is a modified version of the template she used in her class.}}
%\address{Department of Computer Science\\ ETH Z\"urich\\Z\"urich, Switzerland}

@@

\cfbolz{citation again needed for the whole subsection}

@@

In this section, we describe the general model of our TM system. This
should clarify the general semantics in commonly used terms from
the literature.

@@

As required for TM systems, we guarantee complete \emph{isolation}
and \emph{atomicity} for transactions at all times. Furthermore,
the isolation provides full \emph{opacity} to always guarantee a consistent
read set.

To support irreversible operations that cannot be undone when we abort
a transaction (e.g. I/O, syscalls, and non-transactional code in
general), we employ \emph{irrevocable} or \emph{inevitable
transactions}. These transactions are always guaranteed to
commit. There is always at most one such transaction running in the
system, thus their execution is serialised. With this guarantee,
providing \emph{strong isolation} and \emph{serializability} between
non-transactional code is possible by making the current transaction
inevitable right before running irreversible operations.

@@

In order to reference these objects, we need references that are valid
in all threads and automatically point to the private copies. Since
an object's offset inside a segment is the same in all segments, we
can use this offset to reference objects. Because all segments are
copies of each other, this \emph{Segment Offset (SO)} points to the
private version of an object in all threads\,/\,segments. To then
translate this SO to a real virtual memory address when used inside a
thread, we need to add the thread's segment start address to the
SO. The result of this operation is called a \emph{Linear Address
  (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.

x86-CPUs provide a feature called \emph{memory segmentation}. It
performs this translation from a SO to a LA directly in hardware.  We
can use the segment register $\%gs$, which is mostly unused in current
applications. When this register points to a thread's segment start
address, we can instruct the CPU to perform the above translation from
a reference of the form $\%gs{::}SO$ to the right LA on its own. This
process is efficient enough that we can do it on every access to an
object.

@@

To provide atomicity for a transaction, we want to make changes globally
visible on commit. We also need to be able to completely abort a
transaction without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronize all threads
  so that all of them are waiting in a safe point. In the committing
  transaction, we go through all objects in the write set and check if
  another transaction in a different segment read the same object.
  Conflicts are resolved again by either the committing or the other
  transaction waiting or aborting.\\
  We then push all changes of modified objects in private pages to all
  the pages in other segments, including the sharing-segment (segment
  0).
\item [{Abort:}] On abort the transaction will forget about all the
  changes it has done. All objects in the write set are reset by
  copying their previous version from the sharing-segment into the
  private pages of the aborting transaction.\\
  Re-sharing these pages instead of resetting the changes would not
  only be slower because it involves a system call, it is also very
  likely the page would get privatised again immediately after re-trying
  the transaction.  Since this again involves a full page copy,
  resetting should be faster than re-sharing.
\end{description}

@@

Garbage collection plays a big role in our TM system. The GC is
generational and has two generations: the \emph{young} and the
\emph{old} generation.

The \textbf{young generation}, where objects are considered to be
\emph{young} and reside in the \emph{Nursery}, is collected by
\emph{minor collections}. These collections move the surviving objects
out of the nursery into the old object space, which can be done
without stopping other threads. This is done either if the nursery has
no space left anymore or if we are committing the current
transaction. Consequently, all objects are old and the nursery empty
after a transaction commits.  Furthermore, all objects in the nursery
were always created in the current transaction. This fact is useful
since we do not need to call any barrier on this kind of objects.

@@

The \textbf{old generation}, where objects are considered to be
\emph{old} and never move again, is collected by \emph{major
  collections}.  These collections are implemented in a stop-the-world
kind of way and first force minor collections in all threads. The
major goal is to free objects in the old objects space. Furthermore,
we optimistically re-share pages that do not need to be private
anymore.

@@

Committing a transaction needs a bit more work. First, we synchronize
all threads so that the committing one is the only one running and all
the others are waiting in a safe point. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads/segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit (see \ref{subsub:contentionmanagement}).

After verifying that there are no conflicts anymore, we copy all our
changes done to the objects in the write set to all other segments,
including the sharing-segment. This is safe since we synchronised all
threads. We also need to push overflow objects generated by minor
collections to other segments, since they may reside partially in
private pages. At that point we also get a new
\lstinline!overflow_number!  by increasing a global one, so that it
stays globally unique for each transaction. Increasing the
\lstinline!read_version!  is then done at the start of a new
transaction.

@@

\subsubsection{Contention Management\label{subsub:contentionmanagement}}
